---
title: "CariEDA"
author: "Cari Comnick"
date: "May 29, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(lme4)
```

States that aren't in data: hawaii (we took out), alaska, delaware, maryland, connecticut, new hampshire, and Massachusetts.

```{r}
setwd("~/Desktop/Advanced Statistical Modeling/Final Project")
bees <- read.csv("bees.csv") #new data without hawaii
bees$nCLOTHIANIDIN <- as.numeric(bees$nCLOTHIANIDIN)
bees$nIMIDACLOPRID <- as.numeric(bees$nIMIDACLOPRID)
bees$nTHIAMETHOXAM <- as.numeric(bees$nTHIAMETHOXAM)
bees$nACETAMIPRID <- as.numeric(bees$nACETAMIPRID)
bees$nTHIACLOPRID <- as.numeric(bees$nTHIACLOPRID)
bees$nAllNeonic <- as.numeric(bees$nAllNeonic)
```

```{r}
#take logs of most
hist(bees$stocks)
hist(bees$priceperlb)
hist(bees$yieldpercol) #fine
hist(bees$nCLOTHIANIDIN)
hist(bees$nIMIDACLOPRID)
range(log(bees$nIMIDACLOPRID))
hist(bees$nTHIAMETHOXAM)
range(log(bees$nTHIAMETHOXAM))
hist(bees$nACETAMIPRID)
range(log(bees$nACETAMIPRID))
hist(bees$nTHIACLOPRID)
range(log(bees$nTHIACLOPRID))
hist(bees$nAllNeonic)
```

```{r}
#interactions with price per lb
colnames(bees)
bees$priceperlbFACTOR <- cut_number(bees$priceperlb, 3)
p <- ggplot(bees)

p + geom_point(aes(x=log(nCLOTHIANIDIN),y=yieldpercol, color = priceperlbFACTOR))+geom_smooth(aes(x=log(nCLOTHIANIDIN),y=yieldpercol, color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #maybe interaction between price per lb and nCLOTHIANIDIN

p + geom_point(aes(x=log(nIMIDACLOPRID),y=yieldpercol, color = priceperlbFACTOR))+geom_smooth(aes(x=log(nIMIDACLOPRID),y=yieldpercol, color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #parallel, so no interaction between pplb and nimidacloprid

p + geom_point(aes(x=log(nTHIAMETHOXAM),y=yieldpercol, color = priceperlbFACTOR))+geom_smooth(aes(x=log(nTHIAMETHOXAM),y=yieldpercol, color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between pplb and nTHIAMETHOXAM

p + geom_point(aes(x=log(nACETAMIPRID),y=yieldpercol, color = priceperlbFACTOR))+geom_smooth(aes(x=log(nACETAMIPRID),y=yieldpercol, color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #maybe interaction between pplb and nACETAMIPRID

p + geom_point(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = priceperlbFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #maybe interaction between pplb and nTHIACLOPRID

p + geom_point(aes(x=log(nAllNeonic),y=yieldpercol, color = priceperlbFACTOR))+geom_smooth(aes(x=log(nAllNeonic),y=yieldpercol, color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #maybe interaction
```

```{r}
#check to see if on the right track
model1 <- glm(yieldpercol ~ priceperlb * (1 + nCLOTHIANIDIN + nIMIDACLOPRID + nTHIAMETHOXAM + nACETAMIPRID + nTHIACLOPRID), data = bees, family = "poisson")
summary(model1) #whaaaaat they match up!

model2 <- glm(yieldpercol ~ priceperlb * nAllNeonic, data = bees, family = "poisson")
summary(model2)
```

```{r}
#interactions among chemicals
bees$nCLOTHIANIDINFACTOR <- cut_number(bees$nCLOTHIANIDIN, 2)
bees$nIMIDACLOPRIDFACTOR <- cut_number(bees$nIMIDACLOPRID, 3)
bees$nTHIAMETHOXAMFACTOR <- cut_number(bees$nTHIAMETHOXAM, 3)
bees$nACETAMIPRIDFACTOR <- cut_number(bees$nACETAMIPRID, 2)
bees$nTHIACLOPRIDFACTOR <- cut_number(bees$nTHIACLOPRID, 2) #won't work

q <- ggplot(bees)

#nclothiandin
q + geom_point(aes(x=log(nCLOTHIANIDIN),y=yieldpercol, color = nIMIDACLOPRIDFACTOR))+geom_smooth(aes(x=log(nCLOTHIANIDIN),y=yieldpercol, color = nIMIDACLOPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nIMIDACLOPRID and nCLOTHIANIDIN

q + geom_point(aes(x=log(nCLOTHIANIDIN),y=yieldpercol, color = nTHIAMETHOXAMFACTOR))+geom_smooth(aes(x=log(nCLOTHIANIDIN),y=yieldpercol, color = nTHIAMETHOXAMFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIAMETHOXAM and nCLOTHIANIDIN

q + geom_point(aes(x=log(nCLOTHIANIDIN),y=yieldpercol, color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(nCLOTHIANIDIN),y=yieldpercol, color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #no interaction between nACETAMIPRID and nCLOTHIANIDIN

#imodacloprid
q + geom_point(aes(x=log(nIMIDACLOPRID),y=yieldpercol, color = nTHIAMETHOXAMFACTOR))+geom_smooth(aes(x=log(nIMIDACLOPRID),y=yieldpercol, color = nTHIAMETHOXAMFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nIMIDACLOPRID and nTHIAMETHOXAM

q + geom_point(aes(x=log(nIMIDACLOPRID),y=yieldpercol, color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(nIMIDACLOPRID),y=yieldpercol, color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #doesn't seem to be interaction between nIMIDACLOPRID and nACETAMIPRID

#nTHIAMETHOXAM
q + geom_point(aes(x=log(nTHIAMETHOXAM),y=yieldpercol, color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(nTHIAMETHOXAM),y=yieldpercol, color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIAMETHOXAM and nACETAMIPRID

#nTHIACLOPRID
q + geom_point(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = nCLOTHIANIDINFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = nCLOTHIANIDINFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIACLOPRID and nCLOTHIANIDIN

q + geom_point(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = nIMIDACLOPRIDFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = nIMIDACLOPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #maybe interaction between nTHIACLOPRID and nIMIDACLOPRID

q + geom_point(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = nTHIAMETHOXAMFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = nTHIAMETHOXAMFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIACLOPRID and nTHIAMETHOXAM

q + geom_point(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIACLOPRID and nACETAMIPRID, but the 0s are very much changing this one

model3 <- glm(yieldpercol ~ (nCLOTHIANIDIN + nIMIDACLOPRID + nTHIAMETHOXAM + nACETAMIPRID + nTHIACLOPRID)^2, data = bees, family = "poisson")
summary(model3) 
#all sig except nCLOTHIANIDIN:nTHIAMETHOXAM, nIMIDACLOPRID:nACETAMIPRID, and nTHIAMETHOXAM:nTHIACLOPRID, kinda matches
```

```{r}
#interactions between stocks and everything
bees$stocksFACTOR <- cut_number(bees$stocks, 3)
r <- ggplot(bees)

r + geom_point(aes(x=log(stocks),y=yieldpercol, color = nIMIDACLOPRIDFACTOR))+geom_smooth(aes(x=log(stocks),y=yieldpercol, color = nIMIDACLOPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and nCLOTHIANIDIN

r + geom_point(aes(x=log(stocks),y=yieldpercol, color = nIMIDACLOPRIDFACTOR))+geom_smooth(aes(x=log(stocks),y=yieldpercol, color = nIMIDACLOPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and nIMIDACLOPRID

r + geom_point(aes(x=log(stocks),y=yieldpercol, color = nTHIAMETHOXAMFACTOR))+geom_smooth(aes(x=log(stocks),y=yieldpercol, color = nTHIAMETHOXAMFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and nTHIAMETHOXAM

r + geom_point(aes(x=log(stocks),y=yieldpercol, color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(stocks),y=yieldpercol, color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and nACETAMIPRID

r + geom_point(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = stocksFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=yieldpercol, color = stocksFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and nTHIACLOPRID (maybe, zeros seem to be weird)

r + geom_point(aes(x=log(stocks),y=yieldpercol, color = priceperlbFACTOR))+geom_smooth(aes(x=log(stocks),y=yieldpercol, color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and and price per lb
```

```{r}
#do we need a random slope as well? looks like it
bees$year <- as.numeric(bees$year)
p2 <- ggplot(bees, aes(x = year, y = yieldpercol, color = state))
p2+ #geom_line() + 
  theme(legend.position = "none") + stat_smooth(method = "lm", se = FALSE)
p2+ geom_line()
p2+theme(legend.position = "none") + stat_smooth(method = "loess", se = FALSE)
```

```{r}
#unconditional means
bees.lmer <- lmer(yieldpercol ~ 1 + (1|state), data = bees)
summary(bees.lmer)

#unconditional growth
bees.lmer2 <- lmer(yieldpercol ~ yrsince1998 + (1|state), data = bees)
summary(bees.lmer2)
```

```{r}
#build a model
bees.lmer3 <- lmer(yieldpercol ~ yrsince1998 + (log(priceperlb) + log(nCLOTHIANIDIN) + log(nIMIDACLOPRID) + log(nTHIAMETHOXAM) + log(nACETAMIPRID) + log(nTHIACLOPRID))*log(stocks) + log(priceperlb):log(nCLOTHIANIDIN) + log(priceperlb):log(nTHIAMETHOXAM) + log(priceperlb):log(nACETAMIPRID)+ log(priceperlb):log(nTHIACLOPRID)+ log(nIMIDACLOPRID):log(nCLOTHIANIDIN) + log(nTHIAMETHOXAM):log(nIMIDACLOPRID) + log(nTHIAMETHOXAM):log(nACETAMIPRID) + log(nTHIACLOPRID):log(nCLOTHIANIDIN)+ log(nTHIACLOPRID):log(nIMIDACLOPRID) + log(nTHIACLOPRID):log(nTHIAMETHOXAM) + log(nTHIACLOPRID):log(nACETAMIPRID)+ (yrsince1998|state), data=bees)
summary(bees.lmer3, cor=FALSE)

bees.lmer4 <- lmer(yieldpercol ~ yrsince1998 + log(nIMIDACLOPRID) + log(nACETAMIPRID) + (log(priceperlb) + log(nCLOTHIANIDIN) + log(nTHIAMETHOXAM) + log(nTHIACLOPRID))*log(stocks) + log(nTHIAMETHOXAM):log(nACETAMIPRID) + log(nTHIACLOPRID):log(nCLOTHIANIDIN)+ log(nTHIACLOPRID):log(nACETAMIPRID)+ (yrsince1998|state), data=bees)
summary(bees.lmer4)

bees.lmer3.1 <- update(bees.lmer3, REML=F)
bees.lmer4.1 <- update(bees.lmer4, REML=F)

D<-2*(logLik(bees.lmer3.1) - logLik(bees.lmer4.1))
1-pchisq(D,10)

anova(bees.lmer3.1,bees.lmer4.1) #check, prefer smaller model

bees.lmer5 <- lmer(yieldpercol ~ yrsince1998 + log(nIMIDACLOPRID) + log(nACETAMIPRID) + log(nCLOTHIANIDIN) + log(nTHIACLOPRID) + (log(priceperlb) + log(nTHIAMETHOXAM))*log(stocks) + log(nTHIACLOPRID):log(nCLOTHIANIDIN)+ (yrsince1998|state), data=bees)
summary(bees.lmer5, cor=FALSE)
bees.lmer5.1 <- update(bees.lmer5, REML=F)

anova(bees.lmer4.1,bees.lmer5.1)  #prefer smaller model

bees.lmer6 <- lmer(yieldpercol ~ yrsince1998 + log(nIMIDACLOPRID) + log(nTHIAMETHOXAM)  + log(priceperlb)*log(stocks) + (yrsince1998|state), data=bees)
summary(bees.lmer6, cor=FALSE)
bees.lmer6.1 <- update(bees.lmer6, REML=F)

anova(bees.lmer5.1,bees.lmer6.1) #prefer smaller model

bees.lmer7 <- lmer(yieldpercol ~ yrsince1998 + log(priceperlb)*log(stocks) + (yrsince1998|state), data=bees)
summary(bees.lmer7, cor=FALSE)
bees.lmer7.1 <- update(bees.lmer7, REML=F)

anova(bees.lmer6.1,bees.lmer7.1) #prefer this model, now all t values two
```

```{r}
hist(log(bees$totalprod)) #looks okay

#unconditional means
bees2.lmer <- lmer(log(totalprod2) ~ 1 + (1|state), data = bees)
summary(bees2.lmer)

#unconditional growth
bees2.lmer2 <- lmer(log(totalprod2) ~ yrsince1998 + (1|state), data = bees)
summary(bees2.lmer2)

#do we need random slope
bees2.lmer2 <- lmer(log(totalprod2) ~ yrsince1998 + (yrsince1998|state), data = bees)
bees2.lmer2b <- lmer(log(totalprod2) ~ yrsince1998 + (1|state), data = bees)
L1 <- logLik(bees2.lmer2)
L0 <- logLik(bees2.lmer2b)
D <- 2*(L1-L0)
.5*(1-pchisq(D, 2)) + .5*(1-pchisq(D,1)) #we do need the random slope

summary(bees2.lmer2)
```

```{r}
hist(bees$totalprod)
bees$totalprod2 <- bees$totalprod/10000 #scale to be smaller
hist(log(bees$totalprod2))

#build a model with log(totalprod) as response
bees2.lmer3 <- lmer(log(totalprod2) ~ (log(numcol) + log(stocks) + log(priceperlb) + yrsince1998 + log(nCLOTHIANIDIN) + log(nIMIDACLOPRID) + log(nTHIAMETHOXAM) + log(nACETAMIPRID) + log(nTHIACLOPRID) + log(nAllNeonic))^2 + (yrsince1998|state), data = bees)
summary(bees2.lmer3, cor = FALSE)

bees2.lmer4 <- update(bees2.lmer3, . ~ . -(log(numcol):log(priceperlb)+log(numcol):log(nIMIDACLOPRID)+log(numcol):log(nTHIAMETHOXAM)+log(numcol):log(nTHIACLOPRID)+log(numcol):log(nTHIACLOPRID)+log(numcol):log(nAllNeonic)+log(stocks):log(priceperlb)+log(stocks):log(nIMIDACLOPRID)+log(stocks):log(nAllNeonic)+log(priceperlb):yrsince1998+log(priceperlb):log(nCLOTHIANIDIN)+log(priceperlb):log(nIMIDACLOPRID)+log(priceperlb):log(nTHIAMETHOXAM)+log(priceperlb):log(nTHIACLOPRID)+log(priceperlb):log(nAllNeonic)+yrsince1998:log(nCLOTHIANIDIN)+yrsince1998:log(nTHIAMETHOXAM)+yrsince1998:log(nTHIACLOPRID)+log(nCLOTHIANIDIN):log(nACETAMIPRID)+log(nCLOTHIANIDIN):log(nTHIACLOPRID)+log(nCLOTHIANIDIN):log(nAllNeonic) +log(nIMIDACLOPRID):log(nTHIAMETHOXAM)+log(nIMIDACLOPRID):log(nTHIACLOPRID)+log(nTHIAMETHOXAM):log(nACETAMIPRID)+log(nTHIAMETHOXAM):log(nAllNeonic)+log(nACETAMIPRID):log(nTHIACLOPRID)+log(nACETAMIPRID):log(nAllNeonic)))

summary(bees2.lmer4, cor = FALSE)

bees2.lmer3.1 <- update(bees2.lmer3, REML=FALSE)
bees2.lmer4.1 <- update(bees2.lmer4, REML=FALSE)
anova(bees2.lmer3.1,bees2.lmer4.1) #prefer smaller model

bees2.lmer5 <- update(bees2.lmer4, . ~ . -(log(numcol):log(nACETAMIPRID)+log(nCLOTHIANIDIN):log(nTHIAMETHOXAM)+log(nTHIAMETHOXAM):log(nTHIACLOPRID)+log(nTHIACLOPRID):log(nAllNeonic)))
summary(bees2.lmer5, cor = FALSE)

bees2.lmer5.1 <- update(bees2.lmer5, REML=FALSE)
anova(bees2.lmer4.1,bees2.lmer5.1) #prefer smaller model

bees2.lmer6 <- update(bees2.lmer5, . ~ . -(log(stocks):log(nTHIAMETHOXAM)+log(stocks):log(nACETAMIPRID)+log(stocks):log(nTHIACLOPRID)+yrsince1998:log(nACETAMIPRID)+yrsince1998:log(nAllNeonic)+log(nIMIDACLOPRID):log(nAllNeonic)))
summary(bees2.lmer6, cor = FALSE)

bees2.lmer6.1 <- update(bees2.lmer6, REML=FALSE)
anova(bees2.lmer5.1,bees2.lmer6.1) #prefer smaller model

bees2.lmer7 <- update(bees2.lmer6,. ~ . -(yrsince1998:log(nIMIDACLOPRID)))
summary(bees2.lmer7, cor = FALSE)

bees2.lmer7.1 <- update(bees2.lmer7, REML=FALSE)
anova(bees2.lmer6.1,bees2.lmer7.1) #prefer smaller model

#start taking out non-interaction terms
bees2.lmer8 <- update(bees2.lmer7,. ~ . -(log(nAllNeonic) + log(nTHIAMETHOXAM) + log(nTHIACLOPRID)))
summary(bees2.lmer8, cor = FALSE)

bees2.lmer8.1 <- update(bees2.lmer8, REML=FALSE)
anova(bees2.lmer7.1,bees2.lmer8.1) #prefer smaller model

bees.lmerFINAL <- bees2.lmer8
<<<<<<< HEAD

#================================
bees.lmer9 <- update(bees2.lmer8,. ~ . -(log(nIMIDACLOPRID):log(nACETAMIPRID)+log(nACETAMIPRID)))

bees.lmer9.1 <- update(bees.lmer9, REML=FALSE)

anova(bees2.lmer8.1,bees2.lmer9.1)
summary(bees.lmerREM2)

bees.lmerREM3 <- update(bees.lmerREM2,. ~ . -(log(nCLOTHIANIDIN):log(nIMIDACLOPRID) + log(nIMIDACLOPRID)))
```
summary(bees.lmerFINAL)


```{r}
predictions <- predict(bees.lmerFINAL, type="response")
bees$predict <- exp(predictions) #pretty okay
```

```{r}
library(HLMdiag)

conditionalResids<-HLMresid(bees.lmerFINAL, level = 1, standardize = T)
plot(fitted(bees.lmerFINAL), conditionalResids)
abline(h=0)
#pretty constant variance


#b
marginalResids <- HLMresid(bees.lmerFINAL, level = "marginal", standardize = T)
plot(fitted(bees.lmerFINAL), marginalResids) #not really any outliers
bees[which(marginalResids < -3),] #looked at smallest ones, seems like they have a lot of values of 1 for different chemicals, overpredicted
#421 546 552
bees[which(marginalResids > 3),] #looked at smallest ones, seems like they have a lot of values of 1 thiac, underpredicted
#331 460

#c
cd <- cooks.distance(bees.lmerFINAL)
plot(cd, type= "h")

dotplot_diag(cd, cutoff = "internal", name = "cooks.distance")
bees[which(cd > 0.02),] #all of them have thiac = 1, 2 in SC
#214 421 545 546 716

#d
lev <- leverage(bees.lmerFINAL, level=1)
<<<<<<< HEAD
=======
plot(lev[,1])
dotplot_diag(lev[,1], cutoff = "internal", name = "leverage")
#547, 543
```


```{r}
#if you take out the two repeated high diagnostic values, can cut 4 terms from the model :O
bees2 <- bees[-c(421,546),]
bees.lmerREM <- lmer(log(totalprod2) ~ log(numcol) + log(stocks) + log(priceperlb) +
    yrsince1998 + log(nCLOTHIANIDIN) + log(nIMIDACLOPRID) + log(nACETAMIPRID) +  
    (yrsince1998 | state) + log(numcol):log(stocks) + log(numcol):yrsince1998 +  
    log(numcol):log(nCLOTHIANIDIN) + log(stocks):yrsince1998 +  
    log(stocks):log(nCLOTHIANIDIN) + log(priceperlb):log(nACETAMIPRID) +  
    log(nCLOTHIANIDIN):log(nIMIDACLOPRID) + log(nIMIDACLOPRID):log(nACETAMIPRID), data = bees2)
summary(bees.lmerREM)

bees.lmerREM2 <- update(bees.lmerREM,. ~ . -(log(nIMIDACLOPRID):log(nACETAMIPRID)))

bees.lmerREM.1 <- update(bees.lmerREM, REML=FALSE)
bees.lmerREM2.1 <- update(bees.lmerREM2, REML=FALSE)

anova(bees.lmerREM.1,bees.lmerREM2.1)
summary(bees.lmerREM2)

bees.lmerREM3 <- update(bees.lmerREM2,. ~ . -(log(nCLOTHIANIDIN):log(nIMIDACLOPRID) + log(nIMIDACLOPRID)))
bees.lmerREM3.1 <- update(bees.lmerREM3, REML=FALSE)

anova(bees.lmerREM2.1,bees.lmerREM3.1)

summary(bees.lmerREM3, cor=FALSE)

bees.lmerFINAL2 <- bees.lmerREM3
```
>>>>>>> b20fcf7ca49e318135ebe6e5a501b0680975eca7

```{r}
conditionalResids<-HLMresid(bees.lmerFINAL2, level = 1, standardize = T)
plot(fitted(bees.lmerFINAL2), conditionalResids)
abline(h=0)
#constant variance


#b
marginalResids <- HLMresid(bees.lmerFINAL2, level = "marginal", standardize = T)
plot(fitted(bees.lmerFINAL2), marginalResids) #not really any outliers
bees[which(marginalResids < -3),] 
#197 214 550
bees[which(marginalResids > 3),] 
#331 459

#c
cd <- cooks.distance(bees.lmerFINAL2)
plot(cd, type= "h")

dotplot_diag(cd, cutoff = "internal", name = "cooks.distance")
bees[which(cd > 0.02),]
#214 334 429 714

#d
lev <- leverage(bees.lmerFINAL2, level=1)
plot(lev[,1])
bees[which(lev[,1] > .3),]
dotplot_diag(lev[,1], cutoff = "internal", name = "leverage")
#542 543 544 545 563


summary(bees.lmerFINAL2)
```

<<<<<<< HEAD
State i, year j
\begin{align*}
Level 1:
\log(TotalProduction_{ij}) = a_i + b_iyrsince1998_{ij} + \gamma \log(numcol)_{ij} + log(stocks) + log(priceperlb) + log(nCLOTHIANIDIN) + log(nACETAMIPRID) + log(numcol):log(stocks) + log(numcol):yrsince1998 + log(numcol):log(nCLOTHIANIDIN) + log(stocks):yrsince1998 + log(stocks):log(nCLOTHIANIDIN) + log(priceperlb):log(nACETAMIPRID) + \epsilon_{ij}

Level 2:
a_i = \alpha + u_i (random int)
b_i = \beta + v_i (random slope)

\end{align*}
 
=======

New EDA for total prod interactions:
```{r}
p <- ggplot(bees)

p + geom_point(aes(x=log(nCLOTHIANIDIN),y=log(totalprod2), color = priceperlbFACTOR))+geom_smooth(aes(x=log(nCLOTHIANIDIN),y=log(totalprod2), color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #since the lines aren't parallel, suggests interaction between price per lb and nCLOTHIANIDIN

p + geom_point(aes(x=log(nIMIDACLOPRID),y=log(totalprod2), color = priceperlbFACTOR))+geom_smooth(aes(x=log(nIMIDACLOPRID),y=log(totalprod2), color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #parallel, so no interaction between pplb and nimidacloprid

p + geom_point(aes(x=log(nTHIAMETHOXAM),y=log(totalprod2), color = priceperlbFACTOR))+geom_smooth(aes(x=log(nTHIAMETHOXAM),y=log(totalprod2), color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between pplb and nTHIAMETHOXAM

p + geom_point(aes(x=log(nACETAMIPRID),y=log(totalprod2), color = priceperlbFACTOR))+geom_smooth(aes(x=log(nACETAMIPRID),y=log(totalprod2), color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between pplb and nACETAMIPRID

p + geom_point(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = priceperlbFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #definitely interaction between pplb and nTHIACLOPRID (? seems like the lots of very small values are really skewing it)

p + geom_point(aes(x=log(nAllNeonic),y=log(totalprod2), color = priceperlbFACTOR))+geom_smooth(aes(x=log(nAllNeonic),y=log(totalprod2), color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #no interaction
```

```{r}
#interactions among chemicals
bees$nCLOTHIANIDINFACTOR <- cut_number(bees$nCLOTHIANIDIN, 2)
bees$nIMIDACLOPRIDFACTOR <- cut_number(bees$nIMIDACLOPRID, 3)
bees$nTHIAMETHOXAMFACTOR <- cut_number(bees$nTHIAMETHOXAM, 3)
bees$nACETAMIPRIDFACTOR <- cut_number(bees$nACETAMIPRID, 2)
bees$nTHIACLOPRIDFACTOR <- cut_number(bees$nTHIACLOPRID, 2)

q <- ggplot(bees)

#nclothiandin
q + geom_point(aes(x=log(nCLOTHIANIDIN),y=log(totalprod2), color = nIMIDACLOPRIDFACTOR))+geom_smooth(aes(x=log(nCLOTHIANIDIN),y=log(totalprod2), color = nIMIDACLOPRIDFACTOR),method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nIMIDACLOPRID and nCLOTHIANIDIN

q + geom_point(aes(x=log(nCLOTHIANIDIN),y=log(totalprod2), color = nTHIAMETHOXAMFACTOR))+geom_smooth(aes(x=log(nCLOTHIANIDIN),y=log(totalprod2), color = nTHIAMETHOXAMFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIAMETHOXAM and nCLOTHIANIDIN

q + geom_point(aes(x=log(nCLOTHIANIDIN),y=log(totalprod2), color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(nCLOTHIANIDIN),y=log(totalprod2), color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #no interaction between nACETAMIPRID and nCLOTHIANIDIN

#imodacloprid
q + geom_point(aes(x=log(nIMIDACLOPRID),y=log(totalprod2), color = nTHIAMETHOXAMFACTOR))+geom_smooth(aes(x=log(nIMIDACLOPRID),y=log(totalprod2), color = nTHIAMETHOXAMFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nIMIDACLOPRID and nTHIAMETHOXAM

q + geom_point(aes(x=log(nIMIDACLOPRID),y=log(totalprod2), color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(nIMIDACLOPRID),y=log(totalprod2), color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nIMIDACLOPRID and nACETAMIPRID

#nTHIAMETHOXAM
q + geom_point(aes(x=log(nTHIAMETHOXAM),y=log(totalprod2), color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(nTHIAMETHOXAM),y=log(totalprod2), color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIAMETHOXAM and nACETAMIPRID

#nTHIACLOPRID
q + geom_point(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = nCLOTHIANIDINFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = nCLOTHIANIDINFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIACLOPRID and nCLOTHIANIDIN

q + geom_point(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = nIMIDACLOPRIDFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = nIMIDACLOPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #maybe interaction between nTHIACLOPRID and nIMIDACLOPRID

q + geom_point(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = nTHIAMETHOXAMFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = nTHIAMETHOXAMFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIACLOPRID and nTHIAMETHOXAM

q + geom_point(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between nTHIACLOPRID and nACETAMIPRID
```

```{r}
#interactions between stocks and everything
bees$stocksFACTOR <- cut_number(bees$stocks, 3)
yrsince1998FACTOR <- cut_number(bees$yrsince1998, 5)
r <- ggplot(bees)

r + geom_point(aes(x=log(stocks),y=log(totalprod2), color = yrsince1998FACTOR))+geom_smooth(aes(x=log(stocks),y=log(totalprod2), color = yrsince1998FACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and years since 1998

r + geom_point(aes(x=log(stocks),y=log(totalprod2), color = nIMIDACLOPRIDFACTOR))+geom_smooth(aes(x=log(stocks),y=log(totalprod2), color = nIMIDACLOPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #maybe interaction between stocks and nCLOTHIANIDIN

r + geom_point(aes(x=log(stocks),y=log(totalprod2), color = nIMIDACLOPRIDFACTOR))+geom_smooth(aes(x=log(stocks),y=log(totalprod2), color = nIMIDACLOPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #no interaction between stocks and nIMIDACLOPRID

r + geom_point(aes(x=log(stocks),y=log(totalprod2), color = nTHIAMETHOXAMFACTOR))+geom_smooth(aes(x=log(stocks),y=log(totalprod2), color = nTHIAMETHOXAMFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and nTHIAMETHOXAM

r + geom_point(aes(x=log(stocks),y=log(totalprod2), color = nACETAMIPRIDFACTOR))+geom_smooth(aes(x=log(stocks),y=log(totalprod2), color = nACETAMIPRIDFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and nACETAMIPRID

r + geom_point(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = stocksFACTOR))+geom_smooth(aes(x=log(nTHIACLOPRID),y=log(totalprod2), color = stocksFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #maybe interaction between stocks and nTHIACLOPRID

r + geom_point(aes(x=log(stocks),y=log(totalprod2), color = priceperlbFACTOR))+geom_smooth(aes(x=log(stocks),y=log(totalprod2), color = priceperlbFACTOR), method = "nls", formula = y ~ a * x + b, se = FALSE, method.args = list(start = list(a = 0.1, b = 0.1))) #interaction between stocks and and price per lb
```


```{r}
bees3.lmer <- lmer(log(totalprod2) ~ (log(priceperlb) + yrsince1998 + log(nCLOTHIANIDIN) + log(nIMIDACLOPRID) + log(nTHIAMETHOXAM) + log(nACETAMIPRID) + log(nTHIACLOPRID))^2 + (yrsince1998|state), data = bees)
summary(bees3.lmer, cor=FALSE)

bees3.lmer1 <- update(bees3.lmer,. ~ . -(log(nCLOTHIANIDIN):log(nIMIDACLOPRID)+log(nIMIDACLOPRID):log(nTHIAMETHOXAM)+log(nIMIDACLOPRID):log(nACETAMIPRID)+log(nIMIDACLOPRID):log(nTHIACLOPRID)+log(nTHIAMETHOXAM):log(nACETAMIPRID)+log(nACETAMIPRID):log(nTHIACLOPRID)))
summary(bees3.lmer1, cor = FALSE)

bees3.lmer.1 <- update(bees3.lmer, REML=F)
bees3.lmer1.1 <- update(bees3.lmer1, REML=F)

anova(bees3.lmer.1,bees3.lmer1.1) #prefer the smaller model


bees3.lmer2 <- update(bees3.lmer1,. ~ . -(log(priceperlb):log(nCLOTHIANIDIN)+log(priceperlb):log(nIMIDACLOPRID)+log(priceperlb):log(nACETAMIPRID)+log(priceperlb):log(nTHIACLOPRID)+yrsince1998:log(nCLOTHIANIDIN)+yrsince1998:log(nACETAMIPRID)+yrsince1998:log(nTHIACLOPRID)))
summary(bees3.lmer2, cor = FALSE)

bees3.lmer2.1 <- update(bees3.lmer2, REML=F)

anova(bees3.lmer1.1,bees3.lmer2.1) #prefer the smaller model



bees3.lmer3 <- update(bees3.lmer2,. ~ . -(log(priceperlb):log(nTHIAMETHOXAM)))
summary(bees3.lmer3, cor= FALSE)

bees3.lmer3.1 <- update(bees3.lmer3, REML=F)

anova(bees3.lmer3.1,bees3.lmer2.1) #prefer the smaller model



bees3.lmer4 <- update(bees3.lmer3,. ~ . -(yrsince1998:log(nIMIDACLOPRID)+yrsince1998:log(nTHIAMETHOXAM)))
summary(bees3.lmer4, cor = FALSE)

bees3.lmer4.1 <- update(bees3.lmer4, REML=F)

anova(bees3.lmer3.1,bees3.lmer4.1) #prefer the smaller model


bees3.lmer5 <- update(bees3.lmer4,. ~ . -(log(nCLOTHIANIDIN):log(nTHIACLOPRID)))
summary(bees3.lmer5, cor = FALSE)

bees3.lmer5.1 <- update(bees3.lmer5, REML=F)

anova(bees3.lmer5.1,bees3.lmer4.1) #prefer the smaller model



bees3.lmer6 <- update(bees3.lmer5,. ~ . -(log(nTHIAMETHOXAM):log(nTHIACLOPRID)+log(nTHIACLOPRID)))
summary(bees3.lmer6, cor = FALSE)

bees3.lmer6.1 <- update(bees3.lmer6, REML=F)

anova(bees3.lmer5.1,bees3.lmer6.1) #prefer the smaller model


bees3.lmer7 <- update(bees3.lmer6,. ~ . -(log(nCLOTHIANIDIN):log(nTHIAMETHOXAM) + log(nTHIAMETHOXAM) +log(nIMIDACLOPRID)))
summary(bees3.lmer7, cor = FALSE) #final

bees3.lmer7 <- lmer(log(totalprod2) ~ log(priceperlb) + yrsince1998 + log(nCLOTHIANIDIN) +  
    log(nACETAMIPRID) + (yrsince1998 | state) + log(priceperlb):yrsince1998 +  
    log(nCLOTHIANIDIN):log(nACETAMIPRID), data = bees)

bees3.lmer7.1 <- update(bees3.lmer7, REML=F)

anova(bees3.lmer7.1,bees3.lmer6.1) #prefer the smaller model
```

```{r}
#interpretation of interactions
#price per lb in a given year
x <- c(seq(0,17,by=1))
2^(-.18)*2^(.016*x)

#confidence intervals
#get the SE of the estimate
#holding year constant, so use interaction and pplb coef
vcov(bees3.lmer7)

0.84485  (1.975189)0.052227
se1 <- sqrt(0.0562002^2+0.0009482^2+2*(1)*(-1)*(-1.295769e-04))
1-(2^(-0.296)*2^(0.0067*x))

1-(2^(-0.0746)*2^(0.0256*x))
```


```{r}
conditionalResids<-HLMresid(bees3.lmer7, level = 1, standardize = T)
plot(fitted(bees3.lmer7), conditionalResids)
abline(h=0)
#pretty constant variance


#b
marginalResids <- HLMresid(bees3.lmer7, level = "marginal", standardize = T)
plot(fitted(bees3.lmer7), marginalResids) #not really any outliers
bees[which(marginalResids < -2.5),] #183 210 266 267 445 544
bees[which(marginalResids > 3),] # 66 459

#c
cd <- cooks.distance(bees3.lmer7)
plot(cd, type= "h")

dotplot_diag(cd, cutoff = "internal", name = "cooks.distance")
bees[which(cd > 0.02),] #260 421

#d
lev <- leverage(bees3.lmer7, level=1)
dotplot_diag(lev[,1], cutoff = "internal", name = "leverage")
#547 543
```

```{r}
confint(bees3.lmer7)
```

```{r}
#interpret pesticide coefficient
#log(nCLOTHIANIDIN) 0.0046102
#log(nACETAMIPRID) 0.0060765
#log(nCLOTHIANIDIN):log(nACETAMIPRID) -0.0026637
#log(y) = 0.0046102*log(cloth)+0.0060765*(log(acet))-0.0026637*log(cloth)*log(acet)
#set cloth=9 and double acet
g <- 1 #doesn't matter what this is
x <- 9 #most common value
logy <- 0.0046102*(x)+0.0060765*(log(g))-0.0026637*x*log(g)
logy1 <- 0.0046102*(x)+0.0060765*(log(2*g))-0.0026637*x*log(2*g)

exp(logy1)/exp(logy)
#0.9876715


#set acet=y and double cloth
y <- 7 #most common value of acet
logy.1 <- 0.0046102*(log(g))+0.0060765*(y)-0.0026637*y*log(g)
logy1.1 <- 0.0046102*(log(2*g))+0.0060765*(y)-0.0026637*y*log(2*g)

exp(logy1.1)/exp(logy.1)
#0.9903184
```


```{r}
for (i in unique(bees$StateName)){
  print(i)
  print(length(which(bees$StateName == i)))
}
#5 states without all 18 observations
```
